Modelling the emergence of whisker barrels

### 2020-07-03 paper/briefpaper/brief.tex
Tweaks from a read through a print out of the paper

% Colours for revisions:
% Custom colours
\definecolor{revgreen}{rgb}{0.220, 0.463, 0.114}
\definecolor{revblue}{rgb}{0.235, 0.471, 0.847}
\definecolor{revred}{rgb}{0.8353, 0.1255, 0.1255}
\definecolor{revorange}{rgb}{0.8235, 0.4275, 0.0}
\definecolor{revgrey}{rgb}{0.6, 0.6, 0.63}
\definecolor{revcomp}{rgb}{0.0, 0.0, 0.0}
\definecolor{revpurp}{rgb}{0.8, 0.1, 0.8}
\definecolor{revpurp2}{rgb}{0.3882, 0.0745, 0.7098}
\definecolor{revpink}{rgb}{0.7098, 0.0745, 0.6431}
\definecolor{revbblack}{rgb}{0, 0, 0.3}
\definecolor{revgblack}{rgb}{0, 0.2, 0}

% Miscellaneous changes
\newcommand{\cmnt}[1]{\textcolor{blue}{#1}}
% Justify gamma/interaction parameter choice. major point 1.
\newcommand{\MPone}[1]{\textcolor{revpink}{#1}}
% Blue changes are major point 2 (competition mechanism) and other, general
% uncategorised changes.
\newcommand{\MPtwo}[1]{\textcolor{revblue}{#1}}
% Parameter search. Major point 3
\newcommand{\MPthreePar}[1]{\textcolor{revbblack}{#1}}
% Sensitivity analysis. Major point 3
\newcommand{\MPthreeSens}[1]{\textcolor{revgblack}{#1}}
% More predictions. Major point 3
\newcommand{\MPthreePred}[1]{\textcolor{revorange}{#1}}
% Justify competition:
\newcommand{\mpfive}[1]{\textcolor{revpurp2}{#1}}
% Minor point 4:
\newcommand{\mpfour}[1]{\textcolor{revgreen}{#1}}
% Minor point 6:
\newcommand{\mpsix}[1]{\textcolor{revpurp}{#1}}
% Describe the metrics
\newcommand{\metrics}[1]{\textcolor{revred}{#1}}

@@

% Note: The key difference between this theory and the chemospecificity theory
% is that of group membership.

@@

% Justify competition term. This pulled from Stuart's Google doc
\mpfive{We reasoned that an arbitrary number of distinct field locations may
  be specified using a minimum of two guidance fields, if the tendency for the
  density of each projection to concentrate is appropriately balanced against
  the tendency of projections that interact more strongly with the guidance
  fields to migrate further in the directions specified by their
  gradients. Accordingly, projections that interact most strongly with a given
  guidance gradient would come to occupy cortical locations at which that
  field has extreme values, leaving adjacent locations available to be
  occupied by projections with the next strongest interactions, and so
  forth. This would in principle allow the \emph{relative} locations of the
  fields to be specified by the relative values of the interaction parameters,
  $\gamma$, and hence for a topological map in the cortex to be specified by a
  spatial ordering of the $\gamma$ values at the level of the thalamus.}

@@

\mpfive{Testing this reasoning required increasing the strength of the
  competition between simulated thalamocortical projections for cortical
  territory, by increasing the tendency for each projection to compete for
  cortical space into which to concentrate. The major modification required
  was thus to introduce into the model an additional source of competition
  between thalamic projections.}
%
The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching,
\MPtwo{whilst ensuring that $a_i$ is conserved over time,} we
redefined
%
\begin{equation} \label{eq:comp}
  \color{revblue}
  \chi_i(\mb{x}, t) = \frac{\epsilon a_i}{N-1} \nabla \textstyle\sum_{j\ne i}^{N}{a_j}.
  \color{black}
\end{equation}
%
\mpfive{This term contributes to the \emph{flux} of axonal branching as an
  additional source of diffusion, scaled by $\epsilon$, which reduces the
  branching density for a given projection where the branches of other
  projections are dense. Note that this operation is local to individual
  afferent projections.}

@@

\MPone{For simplicity, the two simulated thalamic interaction gradients, as
  well as the two cortical guidance gradients, were initially chosen to be
  linear and orthogonal. Hence a given pair of $\gamma$ values corresponds to
  the coordinate of a barreloid center in the VPM. Coordinates, in a reference
  plane defined by the anterior-posterior and medial-lateral axes, were
  estimated from Fig 5d of} \citep{haidarliu_size_2001}, \MPone{and scaled
  such that $\gamma\in\pm 2$. Note that this scaling is arbitrary because
  according to the model the coordinates provide relative position information
  only.}

A cortical boundary enclosing \cmnt{the} 41 \cmnt{macrovibrissae} barrels was
traced from a cytochrome oxidase stain from \cite{zheng_signal_2001}
\cmnt{(using original data kindly supplied by the authors)}, and
Eqs.~\ref{eq:dc}--\ref{eq:comp} were solved for $N=41$ projections on the
resulting domain, $S$, using $M=2$ linear signalling gradients aligned with
the anterior-posterior and medial-lateral axes.  \mpfour{These gradients are
  shown with the barrel field boundary in Fig.\,\ref{fig:main}A for clarity,
  though like ephrin-5 they may be thought of as extending across the cortical
  hemisphere \citep{miller_epha7-ephrin-a5_2006}. Simulations were stepped
  through 30000 iterations of Eqs.\,\ref{eq:dc}--\ref{eq:comp} ($\delta
  t=0.0001$)}

@@

% New section to describe and introduce the metrics
\metrics{First, we used an algorithm introduced by Honda to measure the
  discrepancy of each barrel shape from a Dirichletform shape}
\citep{honda_geometrical_1983}.  \metrics{Low overall values of this
  \emph{Honda-}$\delta$ metric obtained from simulated barrels indicate a
  close correspondence of the simulated barrel field with a Voronoi
  tesselation, and thus with a biological barrel field. For the tesselation
  that is overlaid on the real barrel field in Fig.\,\ref{fig:main}A,
  $\delta=0.025$, and a reduction in $\delta$ in the example simulation over
  time confirmed that an equivalent `good' Voronoi pattern} (see
\citealp{senft_mouse_1991}) \metrics{can emerge within $\approx$~20000
  iterations (Fig.\,\ref{fig:main}D, red circles). Second, we devised a
  \emph{pattern difference} measure that is sensitive to deviations in the
  component shapes and overall topographic registration between two
  tesselations, $\eta$, and we used this measure to compare the simulated
  barrel fields to the real barrel field from which the boundary shape applied
  to the simulation was obtained (see \emph{Materials and Methods} for
  details). A similar reduction in $\eta$ in the development of the example
  simulation confirmed that the shapes and arrangement of emergent connection
  fields came to match those of the real barrel field by around 20000
  iterations (Fig.\,\ref{fig:main}D, black squares). Third, we measured the
  \emph{connection selectivity} at each location on the cortical sheet, as the
  connection density of the most dense projection divided by the sum over all
  projection densities, $\omega$. The connection selectivity increased as the
  barrel map self-organised in the example simulation (Fig.\,\ref{fig:main}D,
  grey hexagons), and the selectivity became concentrated in regions
  overlapping with the emergent barrel centers (Fig.\,\ref{fig:main}E).}

@@

\MPthreePar{The parameters of the example simulation are indicated in
  Fig.\,\ref{fig:paramsweep}A using an asterisk. In
  Fig.\,\ref{fig:paramsweep}B, we also present examples of alternative
  patterns that emerge for different choices of $D$. Decreasing the rate of
  diffusion may be considered equivalent to increasing the overall size of the
  domain, $S$, hence insights into barrel development in species with
  substantially larger barrel fields, such as capybara which do not have
  clearly identifiable barrels, may be gained by studying pattern formation
  when $D$ is small. In this context, it is interesting to note that for small
  $D$, the organization is predicted to be topological but highly irregular,
  with a general expansion in the territory occupied by the central versus
  peripheral domains that would presumably manifest as an absence of
  identifiable barrel fields (Fig.\,\ref{fig:paramsweep}Biii).}

@@

\MPthreeSens{Using the parameters of the example simulation
  (Fig.\,\ref{fig:main}C) we established baseline mean and standard deviations
  from ten independent simulations with initial uniform random values for
  $a(\mb{x},0)\in(0.2,0.4)$, to be $\delta=0.089\pm 0.004$, $\eta=0.2108\pm
  0.002$, and $\omega=0.2165\pm 0.0001$. Repeating with the variation in the
  initial noise doubled ($a(\mb{x},0)\in(0.1,0.5)$), or removed altogether
  ($a(\mb{x},0)=0.3$), generated distributions of $\delta$, $\eta$, and
  $\omega$ that were not statistically different, as established using paired
  two-sample t-tests. Adding noise to the interaction parameters ($\gamma$)
  affected neither the \emph{Honda-}$delta$ or the \emph{connection
    selectivity} measures substantially (see Fig\,\ref{fig:sens}A), and an
  increase in the \emph{pattern difference} reflected an increase in the
  occurrence of topological defects only when perturbations become so large as
  to cause the ordering of $\gamma$ values from neighbouring thalamic sites to
  be switched (see example map Fig\,\ref{fig:sens}Bi). Adding noise to the
  cortical guidance field values, $\rho_1(\mathbf{x})$ and
  $\rho_2(\mathbf{x})$, disrupted pattern formation only for high levels of
  noise applied at short length scales, which manifested as non-straight edges
  at the domain boundaries (Fig\,\ref{fig:sens}Bii). Varying the slope of one
  linear gradient $\rho_1(\mathbf{x})$ while keeping that of the other
  constant caused elongation of the emergent domains along the corresponding
  axis (Fig\,\ref{fig:sens}Biii), while pattern formation was not strongly
  influenced by relaxing the assumption that the gradients of the two cortical
  guidance fields are orthogonal (Fig\,\ref{fig:sens}Biv). Overall, the
  sensitivity analysis revealed that self-organization of barrel-like fields
  in the model is highly robust to a wide range of sources of perturbation.}

% This should be a little less terse than in the original.
\MPthreePred{To further investigate the interplay of genetic and intrinsic
  mechanisms we simulated two well known experimental manipulations of the
  whisker barrel system. First, we simulated a seminal barrel duplication
  experimental paradigm}
\citep{shimogori_fibroblast_2005,assimacopoulos_fibroblast_2012}
\MPthreePred{in which the growth factor Fgf8, which is normally expressed at
  the anterior end of the cortical subplate from around E9.5}
\citep{crossley_mouse_1995}, \MPthreePred{is ectopically expressed (by
  electroporation) also at the posterior pole of a cortical
  hemisphere. Because Fgf8 sets off a chain of gene interactions involved in
  controlling the graded expression of ephrin and cadherin molecules, its
  misexpression at the posterior pole could be expected to mirror those
  gradients which usually vary along the rostrocaudal axis. We assume that
  this results in a mirror of the primary barrel cortex boundary along the
  rostrocaudal axis, to match the experimental results of}
\citealp{assimacopoulos_fibroblast_2012}. \MPthreePred{We also assume that a
  further consequence of the posterior Fgf8 electroporation is the mirrored
  anterior-posterior guidance gradient $\rho_1$ (Fig.\,\ref{fig:fgf8}A). With
  the modified $\rho_1$ guidance gradient the simulation was run with $N=41$
  thalamocortical axon types, using the parameters of the example
  simulation. The result, after 30000 iterations, was two mirror-symmetrical
  barrel fields comprising $2N$ barrels (Fig\,\ref{fig:main}B), consistent
  with the outcome of the original experiments.}

% This figure is text width
\begin{center}
  \includegraphics[width=\linewidth]{./Fig4.png} \captionof{figure}{
    \MPthreePred{Two simulated whisker barrel manipulations.  Guidance fields
      (\textbf{A}) and emergent barrel pattern (\textbf{B}) in a Fgf8
      misexpression experiment
      (c.f.~\citealp{assimacopoulos_fibroblast_2012}), simulated by reflecting
      $\rho_1$ from Fig.\,\ref{fig:main}A at the join of the original boundary
      with its mirror. The model parameters match those in
      Fig.\,\ref{fig:main}C.  Scale bars are 1\,mm. \textbf{C} Simulating
      whisker trimming by a reduction of the competition parameter,
      $\epsilon$, in the expression $\chi_i$ (Eq.\,\ref{eq:comp}) for the C3
      thalamocortical projection. For C3, $\epsilon$ is replaced by
      $m\,\epsilon$, where $m$ is a scalar multiplier. For the rest of the
      projections, $\chi_i$ is left unchanged and all other parameters are
      identical to those reported in Fig.\,\ref{fig:main}C. The pattern formed
      after 30000 steps with $m=0.64$. Barrel C3 is reduced in size as its
      axon branching competes weakly with its neighbours, whose area is
      slightly increased. \textbf{D} Individual barrel area (mm$^2$) plotted
      vs.~$m$ for C3 (black squares) shows that as $m$ is reduced, its area
      goes down, disappearing for $m\approx0.5$, whereas the mean area of
      neighbouring barrels B3, C2, C4, D2 \& D3 (grey circles) increases. The
      dotted grey line indicates $m=0.64$ for comparison with panel C.}}
  \label{fig:fgf8}
\end{center}

\MPthreePred{To investigate the response of the model to environmental
  manipulation, we simulated a whisker deprivation experiment. In the first
  few days after birth, the upper region of the rat's cortex develops into
  distinct layers I--IV (layers V and VI are already distinct at
  birth). During this critical period, removal of the mystacial whiskers by
  electrocauterization, plucking or trimming leads to observable changes in
  brain structures, including the barrel field in cortical layer
  IV~\citep{jeanmonod_mouse_1981}. Amoungst other changes, barrels associated
  with deprived whiskers have a smaller area. This is found whether the animal
  is deprived of an individual whisker~\citep{kossut_effects_1992}, or an
  entire row~\citep{loos_somatosensory_1973}, and demonstrates that
  thalamocortical activity contributes to the developmental process. We
  simulated the trimming of the individual whisker C3 during this critical
  period by reducing the amount by which its thalamocortical projections
  compete for branching. This was achieved by reducing the value of $\epsilon$
  in Eq.\,\ref{eq:comp} for the C3 projection, leaving other simulation
  parameters unchanged. The reduced competition for branching limited the area
  over which the C3 projection could make connections and its resulting field
  size was diminished (Fig.\,\ref{fig:fgf8}C). Emergent fields representing
  the neighbours of the C3 projection, whose ability to compete for branching
  was unchanged, increased in size. The C3 barrel disappeared altogether if
  its competition parameter was reduced below a half of the original value. A
  reduction in the area of barrel C3, comparable that induced by
  \citep{kossut_effects_1992} (65\%) was obtained in simulation when the
  competitiveness of projection C3, $\epsilon$, was reduced to 86\%, as
  indicated in (Fig.\,\ref{fig:fgf8}D).}

@@

\cmnt{In reality, scores of smaller Dirichletform barrels representing the
  microvibrissae form to the right of where the E-row barrels emerge in our
  simulations, presumably via the same competitive processes. Enforcing here
  the same boundary condition as used to represent the true edges of the
  barrel field is necessary to ensure the stability of the simulation, though
  we acknowledge that this region of the boundary is enforced primarily to
  keep the number of simulated projections, and hence the overall
  computational complexity of the simulation, manageable (simulating each
  additional projection introduces an extra \textbf{YYY} new dynamical
  variables).}

@@

The most involved part of this computation is to find the divergence of the
flux of axonal branching, $\mb{J}_i(\mb{x},t)$, the term in parentheses in
Eq.\,\ref{eq:da}:
%
\begin{equation}
  \label{eq:divJ}
  \nabla\vcdot\mb{J}_i(\mb{x},t) = \nabla\vcdot\left(D \nabla
  a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x}) \color{revblue}+
  \frac{\epsilon a_i}{N-1} \nabla \hat{a}_i \color{black}\right),
\end{equation}
%
\MPtwo{where $\hat{a}_i\equiv\sum_{j\ne i}^{N}a_j$.} Note that the sum of the
guidance gradients is time-independent and define $\mb{g}_i(\mb{x}) \equiv
\sum_{j=1}^{M} \gamma_{i,j} \nabla\rho_j(\mb{x})$.  Because the divergence
operator is distributive, Eq.\,\ref{eq:divJ} can be expanded using vector
calculus identities (dropping references to $\mb{x}$ and $t$ for clarity):
%
\begin{equation}
\nabla\vcdot\mb{J}_i = \nabla\vcdot\big(D \nabla a_i\big) -
\nabla\vcdot\big(a_i \mb{g}_i\big) \color{revblue} + \frac{\epsilon}{N-1}\nabla\vcdot\big({a}_i\nabla\hat{a}_i\big)\color{black}.
\end{equation}
%
Applying the vector calculus product rule identity yields
%
\begin{equation} \label{eq:divJExpanded}
%
\nabla\vcdot\mb{J}_i =
%
D \dvrg a_i % term1
-
a_i\nabla\vcdot\mb{g}_i % term2
-
\mb{g}_i\vcdot\nabla a_i % term3
\color{revblue}
+ \frac{\epsilon a_i}{N-1} \dvrg \hat{a}_i % term1_1
+ \frac{\epsilon}{N-1} \nabla \hat{a}_i\vcdot\nabla{a}_i% term1_2
\color{black}
,
\end{equation}
%
which has \MPtwo{five} elements to compute: i)~$D \dvrg a_i$ (the Laplacian of
$a_i$); ii)~a time-independent modulator of $a_i$ (because
$\nabla\vcdot\mb{g}_i$ is a time-independent static field); iii)~the scalar
product of the static vector field $\mb{g}_i$ and the gradient of $a_i$;
\MPtwo{iv)~the Laplacian of $\hat{a}_i$; and v)~a term involving the gradients
  of $a_i$ and $\hat{a}_i$}. Each of the divergences can be simplified by
means of Gauss's Theorem following \cite{lee_hexagonal_2014}.

@@

% NOTES

%\MPtwo{In other words, we simply scaled the barreloid coordinates in an anterior-posterior/medial-lateral coordinate system to create the $\gamma$ values in the range [-2,2]. Note that positive \emph{and} negative $\gamma$ values were generated; the origin of the coordinate system was located at the approximate center of the thalamic barreloid field. This prevented an overall drift of the cortical pattern away from the centroid of the barrel field and may explain why coordinate-giving transcription factors and the ephrins are often found as pairs acting in opposing directions.}  \MPtwo{We concentrated on macrovibrissae (excluding microvibrissae) because it is easy to identify their corresponding structures both in the thalamus and in the cortex. We also excluded the 5 rhinal whiskers because their cortical representation is in a separate island, distinct, and somewhat removed from the representation of the macrovibirissae.}}

%\MPtwo{Thus, we suggest that there exist a pair of signalling molecules whose expression varies linearly (with orthogonal gradients) across the thalamus, analgous to the ephrins, whose existence is well known in the cortex. Thalamocortical axons growing from a region in the thalamus with high expression of one of these molecules will interact with a corresponding cortical signalling molecule more strongly than an axon emanating from a region with low expression of the thalamic signal.}

>>>>

\input{sebcolour.tex}

% Miscellaneous changes
\newcommand{\cmnt}[1]{\textcolor{colcmnt}{#1}}
% Justify gamma/interaction parameter choice. major point 1.
\newcommand{\MPone}[1]{\textcolor{colmpone}{#1}}
% Blue changes are major point 2 (competition mechanism) and other, general
% uncategorised changes.
\newcommand{\MPtwo}[1]{\textcolor{colmptwo}{#1}}
% Parameter search. Major point 3
\newcommand{\MPthreePar}[1]{\textcolor{colmpthreepar}{#1}}
% Sensitivity analysis. Major point 3
\newcommand{\MPthreeSens}[1]{\textcolor{colmpthreesens}{#1}}
% More predictions. Major point 3
\newcommand{\MPthreePred}[1]{\textcolor{colmpthreepred}{#1}}
% Justify competition:
\newcommand{\mpfive}[1]{\textcolor{colmpfive}{#1}}
% Minor point 4:
\newcommand{\mpfour}[1]{\textcolor{colmpfour}{#1}}
% Minor point 6:
\newcommand{\mpsix}[1]{\textcolor{colmpsix}{#1}}
% Describe the metrics
\newcommand{\metrics}[1]{\textcolor{colmetrics}{#1}}

@@

@@

\mpfive{We reasoned that an arbitrary number of distinct field locations may
  be specified using a minimum of two guidance fields, if the tendency for the
  density of each projection to concentrate is appropriately balanced against
  the tendency of projections that interact more strongly with the guidance
  fields to migrate further in the directions specified by their
  gradients. Accordingly, projections that interact most strongly with a given
  guidance gradient would come to occupy cortical locations at which that
  field has extreme values, leaving adjacent locations available to be
  occupied by projections with the next strongest interactions, and so
  forth. This would in principle allow the \emph{relative} locations of the
  fields to be specified by the relative values of the interaction parameters,
  $\gamma$, and hence for a topological map in the cortex to be specified by a
  spatial ordering of the $\gamma$ values at the level of the thalamus.}

@@

\mpfive{Testing this reasoning required increasing the strength of the
  competition between simulated thalamocortical projections for cortical
  territory, by increasing the tendency for each projection to compete for
  cortical space into which to concentrate. The major modification required
  was thus to introduce into the model an additional source of competition
  between thalamic projections.}
%
The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching,
\MPtwo{whilst ensuring that $a_i$ is conserved over time,} we
redefined
%
\begin{equation} \label{eq:comp}
  \color{colmptwo}
  \chi_i(\mb{x}, t) = \frac{\epsilon a_i}{N-1} \nabla \textstyle\sum_{j\ne i}^{N}{a_j}.
  \color{black}
\end{equation}
%
\mpfive{This term contributes to the \emph{flux} of axonal branching as an
  additional source of diffusion, scaled by $\epsilon$, which reduces the
  branching density for a given projection where the branches of other
  projections are dense. Note that this operation is local to individual
  afferent projections.}

@@

\MPone{For simplicity, the two simulated thalamic interaction gradients, as
  well as the two cortical guidance gradients, were initially chosen to be
  linear and orthogonal. Hence a given pair of $\gamma$ values corresponds to
  the coordinate of a barreloid center in the VPM. Coordinates, in a reference
  plane defined by the anterior-posterior and medial-lateral axes, were
  estimated from Fig 5d of} \cite{haidarliu_size_2001}, \MPone{and scaled
  such that $\gamma\in\pm 2$. Note that this scaling is arbitrary because
  according to the model the coordinates provide relative position information
  only.}

A cortical boundary enclosing \cmnt{the} 41 \cmnt{macrovibrissae} barrels was
traced from a cytochrome oxidase stain from \cite{zheng_signal_2001}
\cmnt{(using original data kindly supplied by the authors)}, and
Eqs.~\ref{eq:dc}--\ref{eq:comp} were solved for $N=41$ projections on the
resulting domain, $S$, using $M=2$ linear signalling gradients aligned with
the anterior-posterior and medial-lateral axes.  \mpfour{These gradients are
  shown with the barrel field boundary in Fig.\,\ref{fig:main}A for clarity,
  though like ephrin-5 they may be thought of as extending across the cortical
  hemisphere \citep{miller_epha7-ephrin-a5_2006}. Simulations were stepped
  through 30000 iterations of Eqs.\,\ref{eq:dc}--\ref{eq:comp} ($\delta
  t=0.0001$).}

@@

% New section to describe and introduce the metrics
\metrics{First, we used an algorithm introduced by Honda to measure the
  discrepancy of each barrel shape from a Dirichletform shape}
\citep{honda_geometrical_1983}.  \metrics{Low overall values of this
  \emph{Honda-}$\delta$ metric obtained from simulated barrels indicate a
  close correspondence of the simulated barrel field with a Voronoi
  tesselation, and thus with a biological barrel field. For the tesselation
  that is overlaid on the real barrel field in Fig.\,\ref{fig:main}A,
  $\delta=0.025$, and a reduction in $\delta$ in the example simulation over
  time confirmed that an equivalent `good' Voronoi pattern} (see
\citealp{senft_mouse_1991}) \metrics{can emerge within $\approx$~20000
  iterations (Fig.\,\ref{fig:main}D, red circles). Second, we devised a
  \emph{pattern difference} measure that is sensitive to deviations in the
  component shapes and overall topographic registration between two
  tesselations, $\eta$, and we used this measure to compare the simulated
  barrel fields to the real barrel field from which the boundary shape applied
  to the simulation was obtained (see \emph{Materials \& Methods} for
  details). A similar reduction in $\eta$ in the development of the example
  simulation confirmed that the shapes and arrangement of emergent connection
  fields came to match those of the real barrel field by around 20000
  iterations (Fig.\,\ref{fig:main}D, black squares). Third, we measured the
  \emph{connection selectivity} at each location on the cortical sheet, as the
  connection density of the most dense projection divided by the sum over all
  projection densities, $\omega$. The connection selectivity increased as the
  barrel map self-organised in the example simulation (Fig.\,\ref{fig:main}D,
  grey hexagons), and the selectivity became concentrated in regions
  overlapping with the emergent barrel centers (Fig.\,\ref{fig:main}E).}

@@

\MPthreePar{The parameters of the example simulation are indicated in
  Fig.\,\ref{fig:paramsweep}A using an asterisk. In
  Fig.\,\ref{fig:paramsweep}B, we also present examples of alternative
  patterns that emerge for different choices of $D$. Decreasing the rate of
  diffusion may be considered equivalent to increasing the overall size of the
  domain, $S$, hence insights into barrel development in species with
  substantially larger barrel fields, such as capybara which do not have
  clearly identifiable barrels, may be gained by studying pattern formation
  when $D$ is small. In this context, it is interesting to note that for small
  $D$, the organization is predicted to be topological but highly irregular,
  with a general expansion in the territory occupied by the central versus
  peripheral domains that would presumably manifest as an absence of
  identifiable barrel fields (Fig.\,\ref{fig:paramsweep}B\,iii).}

@@

\MPthreeSens{Using the parameters of the example simulation
  (Fig.\,\ref{fig:main}C) we established baseline mean and standard deviations
  from ten independent simulations with initial uniform random values for
  $a(\mb{x},0)\in(0.2,0.4)$, to be $\delta=0.089\pm 0.004$, $\eta=0.2108\pm
  0.002$, and $\omega=0.2165\pm 0.0001$. Repeating with the variation in the
  initial noise doubled ($a(\mb{x},0)\in(0.1,0.5)$), or removed altogether
  ($a(\mb{x},0)=0.3$), generated distributions of $\delta$, $\eta$, and
  $\omega$ that were not statistically different, as established using paired
  two-sample t-tests. Adding noise to the interaction parameters ($\gamma$)
  affected neither the \emph{Honda-}$\delta$ or the \emph{connection
    selectivity} measures substantially (see Fig.\,\ref{fig:sens}A), and an
  increase in the \emph{pattern difference} reflected an increase in the
  occurrence of topological defects only when perturbations become so large as
  to cause the ordering of $\gamma$ values from neighbouring thalamic sites to
  be switched (see example map Fig.\,\ref{fig:sens}B\,i). Adding noise to the
  cortical guidance field values, $\rho_1(\mathbf{x})$ and
  $\rho_2(\mathbf{x})$, disrupted pattern formation only for high levels of
  noise applied at short length scales, which manifested as non-straight edges
  at the domain boundaries (Fig.\,\ref{fig:sens}B\,ii). Varying the slope of one
  linear gradient $\rho_1(\mathbf{x})$ while keeping that of the other
  constant caused elongation of the emergent domains along the corresponding
  axis (Fig.\,\ref{fig:sens}B\,iii), while pattern formation was not strongly
  influenced by relaxing the assumption that the gradients of the two cortical
  guidance fields are orthogonal (Fig.\,\ref{fig:sens}B\,iv). Overall, the
  sensitivity analysis revealed that self-organization of barrel-like fields
  in the model is highly robust to a wide range of sources of perturbation.}

\MPthreePred{To further investigate the interplay of genetic and intrinsic
  mechanisms we simulated two well known experimental manipulations of the
  whisker barrel system. First, we simulated a seminal barrel duplication
  experimental paradigm}
\citep{shimogori_fibroblast_2005,assimacopoulos_fibroblast_2012}
\MPthreePred{in which the growth factor Fgf8, which is normally expressed at
  the anterior end of the cortical subplate from around E9.5}
\citep{crossley_mouse_1995}, \MPthreePred{is ectopically expressed (by
  electroporation) also at the posterior pole of a cortical
  hemisphere. Because Fgf8 sets off a chain of gene interactions involved in
  controlling the graded expression of ephrin and cadherin molecules, its
  misexpression at the posterior pole could be expected to mirror those
  gradients which usually vary along the rostrocaudal axis. We assume that
  this results in a mirror of the primary barrel cortex boundary along the
  rostrocaudal axis, to match the experimental results of}
\cite{assimacopoulos_fibroblast_2012}. \MPthreePred{We also assume that a
  further consequence of the posterior Fgf8 electroporation is the mirrored
  anterior-posterior guidance gradient $\rho_1$ (Fig.\,\ref{fig:fgf8}A). With
  the modified $\rho_1$ guidance gradient the simulation was run with $N=41$
  thalamocortical axon types, using the parameters of the example
  simulation. The result, after 30000 iterations, was two mirror-symmetrical
  barrel fields comprising $2N$ barrels (Fig.\,\ref{fig:fgf8}B), consistent
  with the outcome of the original experiments.}

% This figure is text width
\begin{center}
  \includegraphics[width=\linewidth]{./Fig4.png} \captionof{figure}{
    \MPthreePred{Two simulated whisker barrel manipulations.  Guidance fields
      (\textbf{A}) and emergent barrel pattern (\textbf{B}) in a Fgf8
      misexpression experiment
      (c.f.~\citealp{assimacopoulos_fibroblast_2012}), simulated by reflecting
      $\rho_1$ from Fig.\,\ref{fig:main}A at the join of the original boundary
      with its mirror. The model parameters match those in
      Fig.\,\ref{fig:main}C.  Scale bars are 1\,mm. \textbf{C} Simulating
      whisker trimming by a reduction of the competition parameter,
      $\epsilon$, in the expression $\chi_i$ (Eq.\,\ref{eq:comp}) for the C3
      thalamocortical projection. For C3, $\epsilon$ is replaced by
      $m\,\epsilon$, where $m$ is a scalar multiplier. For the rest of the
      projections, $\chi_i$ is left unchanged and all other parameters are
      identical to those reported in Fig.\,\ref{fig:main}C. The pattern formed
      after 30000 steps with $m=0.86$. As its axon branching competes weakly
      with its neighbours, field C3 is reduced to 65\% of its original size,
      matching the average barrel area reduction observed by }
    \citealp{kossut_effects_1992}. \MPthreePred{\textbf{D} Individual barrel
      area (mm$^2$) plotted vs.~$m$ for C3 (black squares) shows that as $m$
      is reduced, its area goes down, disappearing for $m\approx0.5$, whereas
      the mean area of neighbouring barrels B3, C2, C4, D2 \& D3 (grey
      circles) increases. The dotted grey line indicates $m=0.86$ for
      comparison with panel C.}}
  \label{fig:fgf8}
\end{center}

\MPthreePred{To investigate the response of the model to environmental
  manipulation, we simulated a whisker deprivation experiment. In the first
  few days after birth, the upper region of the rat's cortex develops into
  distinct layers I--IV (layers V and VI are already distinct at
  birth). During this critical period, removal of the mystacial whiskers by
  electrocauterization, plucking or trimming leads to observable changes in
  brain structures, including the barrel field in cortical layer
  IV~\citep{jeanmonod_mouse_1981}. Amoungst other changes, barrels associated
  with deprived whiskers have a smaller area. This is found whether the animal
  is deprived of an individual whisker~\citep{kossut_effects_1992}, or an
  entire row~\citep{loos_somatosensory_1973}, and demonstrates that
  thalamocortical activity contributes to the developmental process. We
  simulated the trimming of the individual whisker C3 during this critical
  period by reducing the amount by which its thalamocortical projections
  compete for branching. This was achieved by reducing the value of $\epsilon$
  in Eq.\,\ref{eq:comp} for the C3 projection, leaving other simulation
  parameters unchanged. The reduced competition for branching limited the area
  over which the C3 projection could make connections and its resulting field
  size was diminished (Fig.\,\ref{fig:fgf8}C). Emergent fields representing
  the neighbours of the C3 projection, whose ability to compete for branching
  was unchanged, increased in size. The C3 barrel disappeared altogether if
  its competition parameter was reduced below a half of the original value. A
  reduction in the area of barrel C3, comparable that induced by
  \citealp{kossut_effects_1992} (65\%) was obtained in simulation when the
  competitiveness of projection C3, $\epsilon$, was reduced to 86\%, as
  indicated in Fig.\,\ref{fig:fgf8}D.}

@@

\cmnt{In reality, scores of smaller Dirichletform barrels representing the
  microvibrissae form to the right of where the E-row barrels emerge in our
  simulations, presumably via the same competitive processes. Enforcing here
  the same boundary condition as used to represent the true edges of the
  barrel field is necessary to ensure the stability of the simulation, though
  we acknowledge that this region of the boundary is enforced primarily to
  keep the number of simulated projections, and hence the overall
  computational complexity of the simulation, manageable (simulating each
  additional projection introduces an extra 13030 new dynamical
  variables).}

@@

The most involved part of this computation is to find the divergence of the
flux of axonal branching, $\mb{J}_i(\mb{x},t)$, the term in parentheses in
Eq.\,\ref{eq:da}:
%
\begin{equation}
  \label{eq:divJ}
  \nabla\vcdot\mb{J}_i(\mb{x},t) = \nabla\vcdot\left(D \nabla
  a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x}) \color{colmptwo}+
  \frac{\epsilon a_i}{N-1} \nabla \hat{a}_i \color{black}\right),
\end{equation}
%
\MPtwo{where $\hat{a}_i\equiv\sum_{j\ne i}^{N}a_j$.} Note that the sum of the
guidance gradients is time-independent and define $\mb{g}_i(\mb{x}) \equiv
\sum_{j=1}^{M} \gamma_{i,j} \nabla\rho_j(\mb{x})$.  Because the divergence
operator is distributive, Eq.\,\ref{eq:divJ} can be expanded using vector
calculus identities (dropping references to $\mb{x}$ and $t$ for clarity):
%
\begin{equation}
\nabla\vcdot\mb{J}_i = \nabla\vcdot\big(D \nabla a_i\big) -
\nabla\vcdot\big(a_i \mb{g}_i\big) \color{colmptwo} + \frac{\epsilon}{N-1}\nabla\vcdot\big({a}_i\nabla\hat{a}_i\big)\color{black}.
\end{equation}
%
Applying the vector calculus product rule identity yields
%
\begin{equation} \label{eq:divJExpanded}
%
\nabla\vcdot\mb{J}_i =
%
D \dvrg a_i % term1
-
a_i\nabla\vcdot\mb{g}_i % term2
-
\mb{g}_i\vcdot\nabla a_i % term3
\color{colmptwo}
+ \frac{\epsilon a_i}{N-1} \dvrg \hat{a}_i % term1_1
+ \frac{\epsilon}{N-1} \nabla \hat{a}_i\vcdot\nabla{a}_i% term1_2
\color{black}
,
\end{equation}
%
which has \MPtwo{five} elements to compute: i)~$D \dvrg a_i$ (the Laplacian of
$a_i$); ii)~a time-independent modulator of $a_i$ (because
$\nabla\vcdot\mb{g}_i$ is a time-independent static field); iii)~the scalar
product of the static vector field $\mb{g}_i$ and the gradient of $a_i$;
\MPtwo{iv)~the Laplacian of $\hat{a}_i$; and v)~a term involving the gradients
  of $a_i$ and $\hat{a}_i$}. Each of the divergences can be simplified by
means of Gauss's Theorem following \cite{lee_hexagonal_2014}.

@@
